---
title: "Outlier detection"
author: "Che-Wei Chang"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Outlier detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!---
The dataset used in this walk through were obtained and pre-processed.
For details, see: http://server228vsl.ipsp.uni-hohenheim.de/CheWeiChang/ggnet/-/blob/master/analysis/data.Rmd

Files for demo were copied from `ggnet` repos and renamed to make them be easier to understand
`gbs_georef_1661inds_beagle_imputed_maf0.01_SNPonly_rmChrUn_LDprune0.1.vcf.gz` 
      -> `IPKbarley_GBS1661landraces_LDpruned.vcf.gz`
`georef_1661ind_geo_coord_for_locator.txt`
      -> `IPK_1661ind_geo_coord.txt`
`alstructure_Q_hat_1661inds.csv`
      -> `IPK_1661inds_alstructure_Q_hat.csv`
      
TO DO:
 - modify the plotting functions to take single input from ggoutlier -> switch plotting modes according to attributes
 - add progress bar to internal functions of ggoutlier
 - make a function to summarize ggoutlier output
--->

## Introduction

In this vignette, we would ...

## Barley landrace data

The data of IPK barley landraces came from the study of [Milner et al. 2019](https://www.nature.com/articles/s41588-018-0266-x).
The SNPs were obtained by aligning GBS sequences against the `Morex v3` genome.
The raw genotypic data of cultivated barley were then filtered with `mac > 10` and `mis < 0.95` and subsequently imputed by `BEAGLE v5.2`.

After imputation, a total of 1661 georeferenced landraces were selected from the whole IPK collection according to the passport data procured from the [IPK BRIDGE database](https://bridge.ipk-gatersleben.de/).
Next, we pruned redundant SNPs based on linkage disequilibrium with `PLINK`.

* 1,661 geo-referenced landraces (no duplicates)
      * 29,219 SNPs; LD prunned with SNP $r^2$ < 0.1; MAF > 0.01; no missing: `IPKbarley_GBS1661landraces_LDpruned.vcf.gz`  

To run `GGoutlieR`, we computed ancestry coefficients for 1,661 geo-referenced landraces using `ALStructure` ([Cabreros and Storey 2019](https://doi.org/10.1534/genetics.119.302159)).
Users can also use other software, such as `STRUCTURE` ([Pritchard et al. 2000](https://doi.org/10.1093/genetics/155.2.945)), `ADMIXTURE` ([Alexander et al. 2009](https://genome.cshlp.org/content/19/9/1655.short)), and `sNMF` ([Frichot et al. 2014](https://doi.org/10.1534/genetics.113.160572)), to infer ancestry coefficients.
The key point of this step is to ensure that your data matrix of ancestry coefficients has individuals in rows and ancestral populations in columns.

The R codes below first read `vcf.gz` with the `vcfR` package and converted `vcf` to a matrix of allele counts.
Next, we used `estimate_d` to identify the optimal number of ancestral populations (`dhat`), and then inferred ancestry coefficients accordingly.
The ancestry coefficients were saved as `IPK_1661inds_alstructure_Q_hat.csv`, which can be found in our repository (`http://server228vsl.ipsp.uni-hohenheim.de/CheWeiChang/ggoutlier/-/tree/master/data/`).

```{r compute_Q, eval=F}
library(vcfR)
# a function to convert VCF to a numberic matrix
vcf_to_nummatrix <- function(vcf){
  require(vcfR)
  out <-
    apply(extract.gt(vcf), 2, function(x){
      c(0,0,1,1,1,2,2)[match(x, c("0/0", "0|0", "0/1", "0|1","1|0", "1/1", "1|1"))]
    })
  return(out)
} # vcf_to_nummatrix end

## read vcf
vcf <- read.vcfR("../data/IPKbarley_GBS1661landraces_LDpruned.vcf.gz")
Xmat <- vcf_to_nummatrix(vcf = vcf) # convert VCF to a numeric matrix
any(is.na(Xmat)) # FALSE

# estimate ancestry coefficients for 1661 individuals
colnames(Xmat) <- gsub(colnames(Xmat), pattern = "^0_", replacement = "") # remove prefix
library(alstructure)
dhat <- estimate_d(Xmat) # optimal d is 5
fit.als <- alstructure(Xmat, d = dhat)
alsQ <- fit.als$Q_hat
colnames(alsQ) <- colnames(Xmat)
write.table(alsQ, "../data/IPK_1661inds_alstructure_Q_hat.csv", row.names = F, col.names = T, sep = ",")
```

Here we used `GGoutlieR` to identify individuals with abnormal geo-genetic associations.


```{r run_ggoutlier, eval=FALSE}
library(GGoutlieR)

geo_coord <- read.table("../data/IPK_1661ind_geo_coord.txt", header = T, stringsAsFactors = F)
rownames(geo_coord) <- geo_coord[,1]
geo_coord <- geo_coord[,-1]
anc.coef <- t(as.matrix(read.csv("../data/IPK_1661inds_alstructure_Q_hat.csv", header = F, stringsAsFactors = F)))
pthres = 0.01

## run GGoutlieR with the approach of genetic space KNN only
geneticknn_res <- ggoutlier(geo_coord = geo_coord,
                            gen_coord = anc.coef,
                            plot_dir = "../fig", 
                            p_thres = pthres, 
                            w_power = 2, 
                            cpu = 4, 
                            klim = c(3,50), 
                            method = "geneticKNN")

## run GGoutlieR with the approach of geographical space KNN only
geoknn_res <- ggoutlier(geo_coord = geo_coord,
                        gen_coord = anc.coef,
                        plot_dir = "../fig", 
                        p_thres = pthres, 
                        w_power = 2, 
                        cpu = 4, 
                        klim = c(3,50), 
                        method = "geoKNN")

## run GGoutlieR with the composite approach
compositeknn_res <- ggoutlier(geo_coord = geo_coord,
                              gen_coord = anc.coef,
                              plot_dir = "../fig", 
                              p_thres = pthres, 
                              w_power = 2, 
                              cpu = 4, 
                              klim = c(3,50), 
                              method = "composite")


```

```{r write_and_read_ggoutlier_output, eval=FALSE, echo=FALSE}

saveRDS(geneticknn_res, file = "../data/IPK_landrace_ggoutlier_geneticKNN_output.RDS")
saveRDS(geoknn_res, file = "../data/IPK_landrace_ggoutlier_geoKNN_output.RDS")
saveRDS(compositeknn_res, file = "../data/IPK_landrace_ggoutlier_composite_output.RDS")

geneticknn_res <- readRDS("../data/IPK_landrace_ggoutlier_geneticKNN_output.RDS")
geoknn_res <- readRDS("../data/IPK_landrace_ggoutlier_geoKNN_output.RDS")
compositeknn_res <- readRDS("../data/IPK_landrace_ggoutlier_composite_output.RDS")



```



```{r show_result_table, eval=TRUE}
head(compositeknn_res$geneticKNN_result$statistics)

```

